From Parametricity to Conservation Laws, via Noether's Theorem

### 2013-07-10 mechanics/paper.tex
Work on the Lagrangian mechanics section

Reynolds' theory of relational parametricity \cite{reynolds82} tells
us that parametrically polymorphic programs automatically satisfy
invariance properties. Such invariance properties are often called
\emph{Free Theorems}, after Wadler \cite{wadler89theorems}, since they
follow ``for free'' from the types of programs, rather than through
detailed study of the program text itself. A classic example,
presented by Wadler, is the free theorem for programs $f$ with the
following type:
\begin{displaymath}
  f : \forall \alpha.~\mathsf{List}~\alpha \to \mathsf{List}~\alpha
\end{displaymath}
Such programs takes lists of $\alpha$s, for any type $\alpha$, to
lists of $\alpha$s. Using Reynolds' theory of relational
parametricity, Wadler showed that \emph{any} $f$ with this type
satisfies the following property:
\begin{displaymath}
  \forall \alpha, \beta, g : \alpha \to \beta, l : \mathsf{List}~\alpha.~\mathrm{map}~g~(f~[\alpha]~l) = f~[\beta]~(\mathrm{map}~g~l)
\end{displaymath}
Thus, any $f$ with the type given above commutes with mapping some
arbitrary function $g$ over lists. Thinking in terms of abstract data
types and change of data representation, this free theorem states that
$f$ is invariant under change of data representation from an arbitrary
type $\alpha$ to another arbitrary type $\beta$, via $g$.

Using invariance under change to derive useful properties is a concept
much older than programming languages. In physics, Noether's theorem
\cite{noether} provides a general way to derive conserved properties
of physical systems from

FIXME: example of translation invariance, with the coupled spring example
\begin{equation}\label{eq:coupled-spring-langrangian}
  L(t,q_1,q_2,\dot{q_1},\dot{q_2}) = \frac{1}{2}m(\dot{q_1}^2 + \dot{q_2}^2) - \frac{1}{2}k(q_1 - q_2)^2
\end{equation}
As a general rule, Lagrangians for classical mechanical systems take
the form $L = T - V$, where $T$ is the kinetic energy of the system
and $V$ is the potential energy. From the Lagrangian, using the
principle of stationary action, we can derive the following two
equations of motion for this system. (We describe the principle of
stationary action, and the process for deriving the equations of
motion fully in Section~\ref{sec:conservation-laws-from-symmetry}.)
For this system, the equations of motion are a pair of ordinary
differential equations (ODEs) that describe how the positions and
velocities of the particles evolve over time:
\begin{displaymath}
  FIXME
\end{displaymath}
We could now proceed to solve these ODEs to further analyse the
behaviour of this system. However, Noether's theorem gives us a
powerful way of gaining insight into properties of these ODEs ``for
free'', without necessarily having to find solutions to them.

The Lagrangian (\ref{eq:coupled-spring-langrangian}) does not refer to
any fixed point in space; only the relative distance between the two
particles, along with their velocities, is relevant. Therefore, the
Langrangian (\ref{eq:coupled-spring-langrangian}) is invariant under
translation in space by some arbitrary displacement $x$:
\begin{equation}\label{eq:coupled-spring-translation-invariance}
  L(t,q_1,q_2,\dot{q_1},\dot{q_2}) = L(t,q_1+x,q_2+x,\dot{q_1},\dot{q_2})
\end{equation}
By Noether's theorem, invariance under spatial translation implies
that the linear momentum of the whole system is constant for all
time. For this system, conservation of linear momentum is stated
mathematically like so:
\begin{equation}\label{eq:coupled-spring-linear-momentum}
  FIXME
\end{equation}
In general, Noether's theorem gives us a way of deriving conserved
properties like (\ref{eq:coupled-spring-linear-momentum}) from
invariance properties like
(\ref{eq:coupled-spring-translation-invariance}). In this case, we
have used invariance under translation in space to derive conservation
of linear momentum. Other common examples include invariance under
translation in time, yielding conservation of energy, and invariance
under rotation, yield conservation of angular momentum. We will see
examples of each of these kinds of invariance, and consequent
conservation laws, in Section~\ref{sec:examples}.

The invariance property stated in
Equation~(\ref{eq:coupled-spring-translation-invariance}) is highly
reminisicent of the

In this paper, we present a way of using a generalised version of
Reynolds' theory of relational parametricity to prove the geometric
invariance properties required for the hypotheses of Noether's
theorem. We build an extension of System F$\omega$ suitable for
writing invariant Lagrangians. In our system, the Lagrangian
(\ref{eq:coupled-spring-langrangian}) we gave above for describing a
system of two particles coupled by a spring will have the type:
\begin{displaymath}
  \forall x \mathord: \mathsf{T}(1).~C^\infty(\mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, x \rangle \times \mathbb{R}\langle 1, x \rangle \times \mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, 0 \rangle, \mathbb{R}\langle 1, 0 \rangle)
\end{displaymath}
where... FIXME

@@

\section{Conservation Laws from Symmetry}
\label{sec:conservation-laws-from-symmetry}

@@

\subsection{The Principle of Stationary Action}

@@

Two points: the Langrangian and generalised coordinates. Do
generalised coordinates as a secondary thing.

@@

\begin{example}
  A single particle under a potential field. Set the potential field
  to be (downward) gravity.
\end{example}

@@

The principle of stationary action

\begin{displaymath}
  \frac{d}{dt}\frac{\partial L}{\partial \dot{q_i}} - \frac{\partial L}{\partial q_i} = 0
\end{displaymath}

Explain the different derivatives going on here

Example: Newton's second law...

\subsection{Noether's Theorem} Noether's theorem provides us with deep
insights into the properties of the solutions of the Euler-Lagrange
equations. FIXME: motivate invariance.

FIXME: Define one-parameter groups

\paragraph{Invariance of the Action} Let
\begin{displaymath}
  \mathcal{S}[q;a;b] = \int_a^b L(t,q(t),\dot{q}(t)) \mathit{dt}
\end{displaymath}
be the action of some physical system described by the Lagrangian
$L$. Assume a differentiable invertible function $\tau : \mathbb{R}
\to \mathbb{R}$ that transforms time in some way and a function
$\sigma : \mathbb{R}^n \to \mathbb{R}^n$ that transforms the vector of
generalised coordinates $\vec{q}$ to another vector of generalised
coordinates $\sigma(\vec{q})$.

We say that the action $\mathcal{S}$ is invariant under the
transformations $\tau$ and $\sigma$ if it is the case that for all
$a$, $b$ and $q$ that:
\begin{equation}\label{eq:invariance}
    \int_a^b L(t,q(t),\dot{q}(t)) \mathit{dt}
    = \int_{\tau(a)}^{\tau(b)} L(s, q^*(s), \dot{q^*}(s)) \mathit{ds}
\end{equation}
where $q^*(s) = (\sigma \circ q \circ \tau^{-1})(s)$ is the path $q$
transformed by $\tau$ and $\sigma$.  FIXME: Remark on the use of $t$
and $s$ in the integrals. By change of variables, the right hand
integral in the above equation is equal to the following integral:
\begin{displaymath}
  \int_a^b L(\tau(t), q^*(\tau(t)), \dot{q^*}(\tau(t)))\cdot \dot{\tau}(t) \mathit{dt}
\end{displaymath}
Since the endpoints $a$ and $b$ are arbitrary

\begin{example}
  Continue a simple example from above to show that its action is
  invariant under (e.g.) time translation.
\end{example}

\paragraph{Noether's Theorem} Noether's theorem applies to actions
that are \emph{continuously} invariant. FIXME: not quite right: write
out the maths first.

(a) parameterise $\tau$ and $\sigma$ by a real value $\epsilon$ such
that we get a continuous symmetry of the action; (b) state Noether's
theorem as we will use it.

\begin{theorem}
  Noether's theorem
\end{theorem}

\begin{example}
  Continue the simple example from above to show that the invariance
  under time translation yields conservation of energy.
\end{example}

State that we will do more examples later on, after we discuss how to
derive symmetry from types.

\section{Symmetry from Types}

@@

\section{Examples}

>>>>

Reynolds' theory of relational parametricity \cite{reynolds82} tells
us that parametrically polymorphic programs automatically satisfy
invariance properties. Such invariance properties are often called
\emph{Free Theorems}, after Wadler \cite{wadler89theorems}, since they
follow ``for free'' from the types of programs, rather than through
detailed study of the program text itself. An illustrative example is
the free theorem for programs $f$ with the following type:
\begin{displaymath}
  f : \forall \alpha.~\mathsf{List}~\alpha \to \mathsf{Nat}
\end{displaymath}
Such programs takes lists of $\alpha$s, for any type $\alpha$, to
natural numbers. Using Reynolds' theory of relational parametricity,
Wadler showed that \emph{any} $f$ with this type satisfies the
following property:
\begin{displaymath}
  \forall \alpha, \beta, g : \alpha \to \beta, l : \mathsf{List}~\alpha.~f~[\alpha]~l = f~[\beta]~(\mathrm{map}~g~l)
\end{displaymath}
Thus, any $f$ with the type given above is invariant under mapping
some arbitrary function $g$ over its input. Thinking in terms of
abstract data types and change of data representation, this free
theorem states that $f$ is invariant under change of data
representation from an arbitrary type $\alpha$ to another arbitrary
type $\beta$, via $g$.

The use of invariance under change to derive useful consequences is a
technique much older than programming languages. In physics,
\emph{Noether's theorem} \cite{noether} provides a general way to
derive conservation laws for physical systems from invariance
properties. Informally speaking, Noether's theorem states that every
continuous symmetry of a physical system implies the existence of a
corresponding conserved quantity. A conserved quantity is a function
of the state of the system that is constant for all time. We
illustrate the use of Noether's theorem with a simple example of a
classical system consisting of two particles of equal mass $m$
connected by a spring with coefficient $k$, such that the particles
are constrained to move one-dimensionally.

Noether's theorem applies to systems described in terms of
\emph{Lagrangians}. For classical mechanics, a Lagrangian is a
function of time, and the positions and velocities of all the
particles in the system, to the difference between the kinetic energy
and the potential energy of the system. For our system with two
particles at positions $x_1$ and $x_2$ with velocities $\dot{x_1}$ and
$\dot{x_2}$ the Lagrangian is:
\begin{equation}\label{eq:coupled-spring-lagrangian}
  L(t,x_1,x_2,\dot{x_1},\dot{x_2}) = \frac{1}{2}m(\dot{x_1}^2 + \dot{x_2}^2) - \frac{1}{2}k(x_1 - x_2)^2
\end{equation}
The first summand represents the total kinetic energy of the system in
terms of the mass and velocities, while the second summand represents
the potential energy contained within the spring due to the distance
between the particles.

From the Lagrangian (\ref{eq:coupled-spring-lagrangian}), using the
\emph{principle of stationary action}, we can derive the following two
equations of motion for this system. (We describe the principle of
stationary action, and the process for deriving the equations of
motion fully in Section~\ref{sec:conservation-laws-from-symmetry}.)
For this system, the equations of motion are a pair of ordinary
differential equations (ODEs) that describe how the positions and
velocities of the particles evolve over time (the second deivative
with respect to time, $\ddot{x_i}$, denotes the acceleration of the
$i$th particle):
\begin{mathpar}
  m\ddot{x_1} = k(x_1 - x_2)

  m\ddot{x_2} = k(x_1 - x_2)
\end{mathpar}
We could now proceed to solve these ODEs to further analyse the
behaviour of this system. However, Noether's theorem gives us a
powerful way of gaining insight into properties of these ODEs ``for
free'', without necessarily having to find solutions to them.

The Lagrangian (\ref{eq:coupled-spring-lagrangian}) does not refer to
any fixed point in space; only the relative distance between the two
particles, along with their velocities, is relevant. Therefore, the
Lagrangian (\ref{eq:coupled-spring-lagrangian}) is invariant under
translation in space by some arbitrary displacement $y$:
\begin{equation}\label{eq:coupled-spring-translation-invariance}
  L(t,x_1,x_2,\dot{x_1},\dot{x_2}) = L(t,x_1+y,x_2+y,\dot{x_1},\dot{x_2})
\end{equation}
By Noether's theorem, invariance under spatial translation implies
that the linear momentum of the whole system is constant for all
time. For this system, conservation of linear momentum is stated
mathematically as a vanishing derivative with respect to time:
\begin{equation}\label{eq:coupled-spring-linear-momentum}
  \frac{d}{\mathit{dt}}~m(\dot{x}_1 + \dot{x_2}) = 0
\end{equation}
In general, Noether's theorem gives us a way of deriving conserved
properties like (\ref{eq:coupled-spring-linear-momentum}) from
invariance properties like
(\ref{eq:coupled-spring-translation-invariance}). In this case, we
have used invariance under translation in space to derive conservation
of linear momentum. Other common examples include invariance under
translation in time, yielding conservation of energy, and invariance
under rotation, yielding conservation of angular momentum. We will see
examples of each of these kinds of invariance, and their consequent
conservation laws, in Section~\ref{sec:examples}.

The invariance property stated in
Equation~(\ref{eq:coupled-spring-translation-invariance}) is highly
reminiscent of the free theorem we stated for the polymorphic program
at the start of this introduction. FIXME

In this paper, we present a way of using a generalised version of
Reynolds' theory of relational parametricity to prove the geometric
invariance properties required for the hypotheses of Noether's
theorem. We build an extension of System F$\omega$ suitable for
writing invariant Lagrangians. In our system, the Lagrangian
(\ref{eq:coupled-spring-lagrangian}) we gave above for describing a
system of two particles coupled by a spring will have the type:
\begin{displaymath}
  \forall x \mathord: \mathsf{T}(1).~C^\infty(\mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, x \rangle \times \mathbb{R}\langle 1, x \rangle \times \mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, 0 \rangle, \mathbb{R}\langle 1, 0 \rangle)
\end{displaymath}
where... FIXME

@@

\section{Conservation Laws from Invariance}
\label{sec:conservation-laws-from-symmetry}

@@

\subsection{Lagrangian Mechanics}

@@

Two points: the Lagrangian and generalised coordinates. Do
generalised coordinates as a secondary thing.

@@

\begin{example}
  We have already seen an example of a Lagrangian, in the
  introduction, describing a system of two particles coupled by a
  spring. Another example of a Lagrangian describes a system
  consisting of a single particle of mass $m$ acted upon by a constant
  downward gravitational field with gravitational acceleration\footnote{is this the right word?} $g$:
  \begin{equation}\label{eq:simple-g-lagrangian}
    L(t,x,y,\dot{x},\dot{y}) = \frac{1}{2}m(\dot{x}^2 + \dot{y}^2) - mgy
  \end{equation}
  Again, the left summand represents the kinetic energy of the system
  in terms of the mass and horizontal and vertical velocities of the
  particle. The right summand represents the potential energy of the
  system due to the gravitational field's strength determined by the
  particle's distance from a fixed baseline at $y = 0$.
\end{example}

@@

The \emph{principle of stationary action} states that the physically
realisable paths $q$ are the paths that yield stationary points (i.e.,
either minima or maxima) of the action $\mathcal{S}[q;a;b]$, for all
$a$ and $b$. Defining what is exactly meant by a stationary point of
the action is beyond the scope of this paper, and can be found in
standard references on the calculus of variations, such as Gelfand and
Fomin \cite{gelfandXXcalculus}. The key point for our purposes is that
a path $q$ satisfies the principle of stationary action if and only if
it satisfies a system of ODEs derived from the Lagrangian $L$, called
the Euler-Lagrange equations:
\begin{equation}\label{eq:euler-lagrange}
  \frac{d}{dt}\frac{\partial L}{\partial \dot{q_i}} - \frac{\partial L}{\partial q_i} = 0
\end{equation}
The ODEs derived from the Lagrangian for a classical mechanical system
are called the \emph{equations of motion} for the system.

The notation in the Euler-Lagrange equation, while standard, is not
exactly clear, especially to a programming language theorist used to a
more careful treatment of variables. The two partial derivatives
$\frac{\partial L}{\partial \dot{q_i}}$ and $\frac{\partial
  L}{\partial q_i}$ take partial derivatives of $L$ with respect to
$\dot{q_i}$ and $q_i$ \emph{as variables}. The derivative with respect
to time, $\frac{d}{dt}$ then treats $\dot{q_i}$ as a function of time,
hence the appearance of the second derivatives $\ddot{q_i}$ in the
equations of motion. Sussman and XXX \cite{sussman} criticise the
standard presentation of Lagrangian mechanics for this kind of
confusing notation, and define a clearer notation based on the Scheme
programming language. We elect to stick with the standard notation as
used in physics textbooks just for reasons of familiarity.

\begin{example}
  Computing the Euler-Lagrange equations for the Lagrangian
  (\ref{eq:simple-g-lagrangian}), above, yields a pair of ODEs, for
  the $x$ and $y$ coordinates of the single particle of the system:
  \begin{mathpar}
    m\ddot{x} = 0

    m\ddot{y} = mg
  \end{mathpar}
  Note that both these ODEs have the form $F = m\ddot{x}$, which is
  exactly Newton's second law. Lagrangian mechanics allows, from the
  principle of stationary action, the derivation of Newton's second
  law, rather than postulating it as an axiom.
\end{example}

\subsection{Noether's Theorem}

As we described in the introduction, Noether's theorem provides us
with deep insights into the properties of the solutions of the
Euler-Lagrange equations. Formally, Noether's theorem requires the
invariance of the action $\mathcal{S}$, rather than invariance of the
Lagrangian itself. We therefore first define what is meant by
invariance of the action, and then state Noether's theorem.

\paragraph{Invariance of the Action} Let
\begin{displaymath}
  \mathcal{S}[q;a;b] = \int_a^b L(t,q(t),\dot{q}(t)) \mathit{dt}
\end{displaymath}
be the action of some physical system described by the Lagrangian
$L$. Assume a differentiable invertible function $\Phi : \mathbb{R}
\to \mathbb{R}$ that transforms time in some way and a function
$\Psi : \mathbb{R}^n \to \mathbb{R}^n$ that transforms the vector of
generalised coordinates $\vec{q}$ to another vector of generalised
coordinates $\Psi(\vec{q})$.

The action $\mathcal{S}$ is \emph{invariant} between the endpoints $a$
and $b$, under the transformations $\Phi$ and $\Psi$, if it is the
case that for all paths $q$:
\begin{equation}\label{eq:invariance}
    \int_a^b L(t,q(t),\dot{q}(t)) \mathit{dt}
    = \int_{\Phi(a)}^{\Phi(b)} L(s, q^*(s), \dot{q^*}(s)) \mathit{ds}
\end{equation}
where $q^*(s) = (\Psi \circ q \circ \Phi^{-1})(s)$ is the path $q$
transformed by $\Phi$ and $\Psi$.  By change of variables, the right
hand integral in the above equation is equal to the following
integral:
\begin{displaymath}
  \int_a^b L(\Phi(t), q^*(\Phi(t)), \dot{q^*}(\Phi(t)))\cdot \dot{\Phi}(t) \mathit{dt}
\end{displaymath}
For our application to classical mechanics, the endpoints $a$ and $b$
are arbitrary, so we can rewrite Equation (\ref{eq:invariance}) to the
following equation between applications of the Lagrangian:
\begin{equation}\label{eq:simplified-invariance}
  L(t,q(t),\dot{q}(t)) = L(\Phi(t), q^*(\Phi(t)), \dot{q^*}(\Phi(t)))\cdot \dot{\Phi}(t)
\end{equation}
For most examples, it will be the case that $\Phi$ has the form
$\Phi(t) = t + t'$ and $\Psi$ is an affine transformation $\Psi(q)
= Gq + x$, where $G$ is an $n\mathord\times n$ real matrix and $x \in
\mathbb{R}^n$. In this case, Equation (\ref{eq:simplified-invariance})
simplifies further, using the definition of $q^*$ and the chain rule
of differentiation, to:
\begin{equation}
  \label{eq:simple-invariance}
  L(t,q(t),\dot{q}(t)) = L(t + t', Gq(t) + x, G\dot{q}(t))
\end{equation}
Equation (\ref{eq:simple-invariance}) is obviously implied by the
following equation, where we treat $q$ and $\dot{q}$ as universally
quantified variables, instead of as functions of time:
\begin{equation}
  \label{eq:simplest-invariance}
  L(t,q,\dot{q}) = L(t + t', Gq + x, G\dot{q})
\end{equation}
Equation (\ref{eq:simplest-invariance}) is the form of invariance that
the free theorems we derive from the types of our typed language for
classical mechanics will usually take. We define our type system for
classical mechanics in Sections~\ref{sec:symmetry-from-types}
and~\ref{sec:types-for-classical-mech} below.

\begin{example}\label{ex:invariance-examples}
  In the introduction, we stated that the Lagrangian
  (\ref{eq:coupled-spring-lagrangian}) describing the spring-coupled
  system is invariant under translation in space. Our running example
  in this section, the Lagrangian (\ref{eq:simple-g-lagrangian}) for
  the single particle under gravity, is not invariant under all
  spatial translations, due to the reference in the potential energy
  term to the distance from the fixed baseline at $y = 0$. It is
  invariant under translations along the $x$-axis, however. Both
  Lagrangians are also invariant in translations in time, a special
  case of Equation~(\ref{eq:simplest-invariance}):
  \begin{displaymath}
    L(t,q,\dot{q}) = L(t+t',q,\dot{q})
  \end{displaymath}
  This invariance property is simply a consequence of the fact that
  neither Lagrangian explicitly mentions the time $t$.
\end{example}

\paragraph{Noether's Theorem} Noether's theorem applies to actions
that are \emph{continuously} invariant. Being continuously invariant
means that, instead of having fixed transformations $\Phi$ and $\Psi$,
we have two families of transformations, $\Phi_\epsilon$ and
$\Psi_\epsilon$, differentiably indexed by a real parameter
$\epsilon$, such that $\Phi_0$ and $\Psi_0$ are the identity function.

\begin{theorem}[Noether]
  If the action
  \begin{displaymath}
    \mathcal{S}[q;a;b] = \int_a^b L(t,q(t),\dot{q}(t))~dt
  \end{displaymath}
  is invariant under $\epsilon$-indexed families of transformations
  $\Phi_\epsilon$ and $\Psi_\epsilon$, then the equation
  \begin{equation}
    \label{eq:conservation-law}
    \frac{d}{dt}\left(\sum_{i=1}^n \frac{\partial L}{\partial \dot{q_i}}\psi_i + \left(L - \sum_{i=1}^n \dot{q_i}\frac{\partial L}{\partial \dot{q_i}}\right) \phi \right) = 0
  \end{equation}
  holds for all paths $q$ satisfying the Euler-Lagrange equations
  (\ref{eq:euler-lagrange}), where $\psi = \left. \frac{\partial
      \Psi}{\partial \epsilon} \right|_{\epsilon = 0}$ and $\phi =
  \left. \frac{\partial \Phi}{\partial \epsilon} \right|_{\epsilon =
    0}$.
\end{theorem}

\begin{proof}
  E.g., Gelfand and Fomin \cite{gelfandXXcalculus}, Section 20.
\end{proof}

\begin{example}
  In Example~\ref{ex:invariance-examples}, we noted that the
  Lagrangians (\ref{eq:coupled-spring-lagrangian}) and
  (\ref{eq:simple-g-lagrangian}) are both invariant under arbitrary
  translations in time. In terms of continuous invariance, their
  corresponding actions are invariant under the families
  $\Phi_\epsilon(t) = t + \epsilon$ and $\Psi_\epsilon(q) =
  q$. Plugging these definitions into
  Equation~(\ref{eq:conservation-law}), we learn that, for these
  systems:
  \begin{displaymath}
    \frac{d}{dt}~\left(L - \sum_{i=1}^2 \dot{q_i}\frac{\partial L}{\partial \dot{q_i}} \right) = 0
  \end{displaymath}
  For the spring-coupled particles Lagrangian,
  (\ref{eq:coupled-spring-lagrangian}), we obtain:
  \begin{displaymath}
    \frac{d}{dt}~\left(\frac{1}{2}m(\dot{x_1}^2 + \dot{x_2}^2) + \frac{1}{2}k(x_1 - x_2) \right) = 0
  \end{displaymath}
  and for the single particle under gravity Lagrangian
  (\ref{eq:simple-g-lagrangian}), we obtain:
  \begin{displaymath}
    \frac{d}{dt}~\left(\frac{1}{2}m(\dot{x}^2 + \dot{y}^2) + mgy\right) = 0
  \end{displaymath}
  In both cases, we have discovered that, as a consequence of
  invariance under translation in time, the total energy of the system
  is conserved (note that the $-$ signs in the original Lagrangians
  have turned into $+$s). In general, invariance under translation in
  time implies, via Noether's theorem, conservation of energy.
\end{example}

This finishes our short introduction to Lagrangian mechanics and
Noether's theorem. FIXME: State that we will do more examples later
on, after we discuss how to derive invariance properties from types.

\section{Invariance from Types}
\label{sec:symmetry-from-types}

@@

\section{Examples}\label{sec:examples}
